Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-18 ebayes_rnaseq_heterosis.tex
updating of shrinkbayes

The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}. A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior. Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})|y)$. To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}. 
Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect. After fitting the model with {\tt ShrinkBayes}, we constructed a measure of heterosis for each gene. For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities. Specifically, the measure of heterosis for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g + \alpha_g > 0 |y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.

@@

Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic
heterosis of traits, such as plant height or grain yield. Because of the unwieldy composite null hypothesis in equation \eqref{e:hypotheses}, existing methods for identifying gene expression heterosis based on RNAseq data are not directly applicable for detecting heterosis genes. \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data. We followed their approach modified slightly to allow for any number of varieties and therefore, possibly, even more complex hypotheses. We presented an empirical Bayes method to obtain MAP estimates for hyperparameters, followed by an MCMC to estimate gene-specific parameters that can be accomplished efficiently via parallelization. This methodology can be conveniently implemented in the statistical software, {\tt Stan}, which also allows for far more complex models. The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.

>>>>

The {\tt ShrinkBayes} package implements empirical Bayesian inference via integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}. A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior. Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})|y)$. To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}. 
Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect. After fitting the model with {\tt ShrinkBayes}, we constructed a measure of heterosis for each gene. For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities. Specifically, the measure of heterosis for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g + \alpha_g > 0 |y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.

@@

Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield. Because of the unwieldy composite null hypothesis in equation \eqref{e:hypotheses}, existing methods for identifying gene expression heterosis based on RNAseq data are not directly applicable for detecting heterosis genes. \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data. We followed their approach, modified slightly to allow for any number of varieties and therefore even more complex hypotheses. We developed an empirical Bayes approach based on obtaining MAP estimates for hyperparameters followed by an MCMC to estimate gene-specific parameters. This methodology can be conveniently implemented in the statistical software, {\tt Stan}. The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.
